---
title: "Batch Effects"
author: "Philipp Sven Lars Schäfer"
date: "`r format(Sys.time(), '%d %B, %Y')`"
editor: source
engine: knitr
---

# Packages

```{r}
suppressPackageStartupMessages({
  library(tidyverse)
  library(ggdark)
  library(vsn)
  library(lme4)
  library(DESeq2)
  library(factoextra)
  library(FactoMineR)
  library(magick) # formatting
})
knitr::knit_hooks$set(crop = knitr::hook_pdfcrop) # formatting

source(file.path("..", "src", "read_data.R"))
source(file.path("..", "src", "generate_targets.R"))
source(file.path("..", "src", "model.R"))
source(file.path("..", "src", "batch_effects.R"))
```

# Data

```{r}
input_dir = file.path("..", "data")
```

```{r}
meta_data <- read_harmonized_meta_data(input_dir)
gene_meta <- read_gene_meta(input_dir)
protein_meta <- read_protein_meta(input_dir)
celltype_meta <- read_celltype_meta(input_dir)

experimental_data <- read_raw_experimental_data(input_dir)
experimental_data <- filter_experimental_data(meta_data=meta_data, 
                                              experimental_data=experimental_data,
                                              gene_meta=gene_meta)

specimen_per_day <- get_specimen_per_day(meta_data=meta_data)
```

# Conclusions

## PBMC Fractions

- For the baseline specimen (day 0), we see a very strong batch effect, because the 2023 dataset is very different

## PBMC Gene Expression

- Very strong cohort-specific effects for baseline specimen and post-booster specimen

## Plasma Antibody Levels

## Plasma Cytokine Concentration by Legendplex

- After removing outlier in the legendplex data...

## Plasma Cytokine Concentration by Olink

- At baseline: About 6.5% of the variance is explained by the dataset, probably also due to 3 outliers from the 2022 datasets. TODO: Investigate this in more detail.

## T Cell Activation

- Specimen are only available for baseline, no cohort-specific effect

## T Cell Polarization

- Specimen are only available for baseline, most of the variance is explained by interaction terms `age_at_boost:dataset` and `infancy_vac:dataset`. TODO: Think about what to make of this.

# Results

```{r}
#| results: asis

##| crop: true # interfers with dynamic toc generation

for (day in names(specimen_per_day)) {
  # day <- names(specimen_per_day)[1]
  
  cat("\n\n")
  cat(paste0("## ", day))
  cat("\n\n")
  
  specimen_df <- specimen_per_day[[day]]
  
  for (assay in names(experimental_data)) {
    # assay <- names(experimental_data)[4]
    
    cat("\n\n")
    cat(paste0("### ", assay))
    cat("\n\n")
    
    feature_col <- experimental_data_settings[[assay]]$feature_col
    value_col <- experimental_data_settings[[assay]]$value_col
    
    assay_df_long <- experimental_data[[assay]]
    #str(assay_df_long)
    
    specimen_df_filtered <- specimen_df %>%
      dplyr::filter(specimen_id %in% unique(assay_df_long$specimen_id))
    
    if (nrow(specimen_df_filtered) <= 3) {
      cat(paste0("Fewer than 3 specimen for this combination of assay and day available"))
    } else {
      assay_mtx_wide <- assay_df_long %>%
        dplyr::select(dplyr::all_of(c("specimen_id", feature_col, value_col))) %>%
        tidyr::pivot_wider(names_from=dplyr::all_of(feature_col), 
                           values_from=dplyr::all_of(value_col)) %>%
        tibble::column_to_rownames(var="specimen_id") %>%
        as.matrix() %>%
        .[as.character(specimen_df_filtered$specimen_id), ]
      
      impute_mode <- "knn"
      stopifnot(impute_mode %in% c("zero", "median_feature", "knn"))
      
      if (any(is.na(assay_mtx_wide))) {
        cat(paste0("Non-zero fraction of NAs: ", 
                   round(mean(is.na(assay_mtx_wide)), 4)))
        # data: matrix with genes in the rows, samples in the columns
        if (impute_mode == "zero") {
          assay_mtx_wide[is.na(assay_mtx_wide)] <- 0
        } else if (impute_mode == "knn") {
          assay_mtx_wide <- t(impute::impute.knn(data=t(assay_mtx_wide))$data)
        } else if (impute_mode == "median_feature") {
          col_medians <- matrixStats::colMedians(assay_mtx_wide, na.rm=TRUE)
          for (colname in colnames(assay_mtx_wide)) {
            col_vec <- assay_mtx_wide[, colname]
            col_vec[is.na(col_vec)] <- col_medians[colname]
            assay_mtx_wide[, colname] <- col_vec
          }
        }
        stopifnot(!any(is.na(assay_mtx_wide)))
      }
      
      if (assay == "pbmc_gene_expression") {
        # normalization using vst from DEseq2
        #str(assay_mtx_wide)
        
        # Create DESeq dataset object
        min_counts <- 200
        min_samples <- 20
      
        genes_to_keep <- colnames(assay_mtx_wide)[
          (colSums(assay_mtx_wide) >= min_counts) &
            (colSums(assay_mtx_wide > 0) >= min_samples)
        ]
        assay_mtx_wide <- assay_mtx_wide[ , genes_to_keep]
        #str(assay_mtx_wide)
        
        # check that all column (gene) names are unique
        stopifnot(length(colnames(assay_mtx_wide)) == length(unique(colnames(assay_mtx_wide))))
        
        assay_mtx_wide <- 
          DESeq2::DESeqDataSetFromMatrix(countData = t(assay_mtx_wide),
                                         colData = tibble::tibble(specimen_id=rownames(assay_mtx_wide)),
                                         design = ~ 1)
        
        # DESeq2 normalization with variance stabilizing transformation (vst)
        assay_mtx_wide <- DESeq2::estimateSizeFactors(assay_mtx_wide, quiet=TRUE) ## estimate size factor
        assay_mtx_wide <- DESeq2::estimateDispersions(assay_mtx_wide, quiet=TRUE)
        assay_mtx_wide <- DESeq2::varianceStabilizingTransformation(assay_mtx_wide, blind=TRUE)
        assay_mtx_wide <- t(assay(assay_mtx_wide))
      }
      
      # dist_mtx <- as.matrix(dist(assay_mtx_wide))
      # dist_mtx_median <- median(dist_mtx[lower.tri(dist_mtx)])
      # dist_mtx_binary <- dist_mtx > (3 * dist_mtx_median)
      # outliers <-
      #  which(rowSums(dist_mtx_binary) > (10*median(rowSums(dist_mtx_binary))))
      # no_outliers <- rownames(dist_mtx)[!rownames(dist_mtx) %in% names(outliers)]
      # assay_mtx_wide <- assay_mtx_wide[no_outliers, ]
      
      #str(assay_mtx_wide)
      suppressMessages({
        suppressWarnings({
          p1 <- pvca_analysis(assay_mtx_wide=assay_mtx_wide,
                              meta_data=meta_data)$p1
        })
      })
      print(p1)
      
      plist <- generate_pca_plots(assay_mtx_wide=assay_mtx_wide,
                                  meta_data=meta_data)
      print(plist$p1)
      print(plist$p2)
      print(plist$p3)
    }
    cat(paste0("\n\n"))
  }
}
```

# Appendix

<button class="accordion-button" type="button" data-bs-toggle="collapse" data-bs-target="#collapseOne" >Session Information</button><div id="collapseOne" class="accordion-collapse collapse"><div>

```{r}
sessionInfo()
```

</div></div>