---
title: "Batch Effects After Normalization"
author: "Philipp Sven Lars Schäfer"
date: "`r format(Sys.time(), '%d %B, %Y')`"
editor: source
engine: knitr
---

# Packages

```{r}
suppressPackageStartupMessages({
  library(tidyverse)
  library(ggdark)
  library(vsn)
  library(lme4)
  library(DESeq2)
  library(factoextra)
  library(FactoMineR)
  library(magick) # formatting
})
knitr::knit_hooks$set(crop = knitr::hook_pdfcrop) # formatting

source(file.path("..", "src", "read_data.R"))
source(file.path("..", "src", "generate_targets.R"))
source(file.path("..", "src", "model.R"))
source(file.path("..", "src", "batch_effects.R"))
source(file.path("..", "src", "normalize_integrate.R"))
```

# Data

```{r}
input_dir = file.path("..", "data")
```

```{r}
celltype_meta <- read_celltype_meta(input_dir)
gene_meta <- read_gene_meta(input_dir)
protein_meta <- read_protein_meta(input_dir)

meta_data <- read_harmonized_meta_data(input_dir)

RECOMPUTE <- TRUE
if (RECOMPUTE) {
  experimental_data <- read_raw_experimental_data(input_dir)
  experimental_data <- filter_experimental_data(meta_data=meta_data, 
                                                experimental_data=experimental_data,
                                                gene_meta=gene_meta)
  experimental_data <- normalize_experimental_data(meta_data=meta_data, 
                                                   raw_experimental_data=experimental_data,
                                                   gene_meta=gene_meta)
  write_rds(experimental_data, 
            file = file.path(input_dir, "prc_datasets", 
                             "normalized_experimental_data.RDS"))
} else {
  experimental_data <- read_rds(file = file.path(input_dir, "prc_datasets", 
                                                 "normalized_experimental_data.RDS"))
}
experimental_data <- experimental_data[-which(names(experimental_data) ==
                                                "pbmc_gene_expression_counts")]
specimen_per_day <- get_specimen_per_day(meta_data=meta_data)
```

# Conclusions

## PBMC Fractions

- No further processing needed I guess.

## PBMC Gene Expression

- Very strong cohort-specific effects for baseline specimen and post-booster specimen

## Plasma Antibody Levels

- Slight batch effect. Not sure whether it needs some kind of batch correction or not.

## Plasma Cytokine Concentration by Legendplex

- Slight batch effect. Not sure whether it needs some kind of batch correction or not. The data distribution looks odd anyway!

## Plasma Cytokine Concentration by Olink

- No further processing needed I guess.

## T Cell Activation

- No further processing needed I guess.

## T Cell Polarization

- Specimen are only available for baseline, most of the variance is explained by interaction terms `age_at_boost:dataset` and `infancy_vac:dataset`. TODO: Think about what to make of this.

# Results

```{r}
#| results: asis

##| crop: true # interfers with dynamic toc generation

for (day in names(specimen_per_day)) {
  # day <- names(specimen_per_day)[1]
  
  cat("\n\n")
  cat(paste0("## ", day))
  cat("\n\n")
  
  specimen_df <- specimen_per_day[[day]]
  
  for (assay in names(experimental_data)) {
    # assay <- names(experimental_data)[4]
    
    cat("\n\n")
    cat(paste0("### ", assay))
    cat("\n\n")
    
    feature_col <- experimental_data_settings[[assay]]$feature_col
    value_col <- experimental_data_settings[[assay]]$value_col
    
    assay_df_long <- experimental_data[[assay]]
    #str(assay_df_long)
    
    specimen_df_filtered <- specimen_df %>%
      dplyr::filter(specimen_id %in% unique(assay_df_long$specimen_id))
    
    if (nrow(specimen_df_filtered) <= 3) {
      cat(paste0("Fewer than 3 specimen for this combination of assay and day available"))
    } else {
      assay_mtx_wide <- assay_df_long %>%
        dplyr::select(dplyr::all_of(c("specimen_id", feature_col, value_col))) %>%
        tidyr::pivot_wider(names_from=dplyr::all_of(feature_col), 
                           values_from=dplyr::all_of(value_col)) %>%
        tibble::column_to_rownames(var="specimen_id") %>%
        as.matrix() %>%
        .[as.character(specimen_df_filtered$specimen_id), ]
      
      impute_mode <- "knn"
      stopifnot(impute_mode %in% c("zero", "median", "knn"))
      
      if (any(is.na(assay_mtx_wide))) {
        cat(paste0("Non-zero fraction of NAs: ", 
                   round(mean(is.na(assay_mtx_wide)), 4)))
        # data: matrix with genes in the rows, samples in the columns
        if (impute_mode == "zero") {
          assay_mtx_wide[is.na(assay_mtx_wide)] <- 0
        } else if (impute_mode == "knn") {
          assay_mtx_wide <- t(impute::impute.knn(data=t(assay_mtx_wide))$data)
        } else if (impute_mode == "median") {
          col_medians <- matrixStats::colMedians(assay_mtx_wide, na.rm=TRUE)
          for (colname in colnames(assay_mtx_wide)) {
            col_vec <- assay_mtx_wide[, colname]
            col_vec[is.na(col_vec)] <- col_medians[colname]
            assay_mtx_wide[, colname] <- col_vec
          }
        }
        stopifnot(!any(is.na(assay_mtx_wide)))
      }
      
      # dist_mtx <- as.matrix(dist(assay_mtx_wide))
      # dist_mtx_median <- median(dist_mtx[lower.tri(dist_mtx)])
      # dist_mtx_binary <- dist_mtx > (3 * dist_mtx_median)
      # outliers <-
      #  which(rowSums(dist_mtx_binary) > (10*median(rowSums(dist_mtx_binary))))
      # no_outliers <- rownames(dist_mtx)[!rownames(dist_mtx) %in% names(outliers)]
      # assay_mtx_wide <- assay_mtx_wide[no_outliers, ]
      
      #str(assay_mtx_wide)
      suppressMessages({
        suppressWarnings({
          p1 <- pvca_analysis(assay_mtx_wide=assay_mtx_wide,
                              meta_data=meta_data)$p1
        })
      })
      print(p1)
      
      plist <- generate_pca_plots(assay_mtx_wide=assay_mtx_wide,
                                  meta_data=meta_data)
      print(plist$p1)
      print(plist$p2)
      print(plist$p3)
    }
    cat(paste0("\n\n"))
  }
}
```

# Appendix

<button class="accordion-button" type="button" data-bs-toggle="collapse" data-bs-target="#collapseOne" >Session Information</button><div id="collapseOne" class="accordion-collapse collapse"><div>

```{r}
sessionInfo()
```

</div></div>

