---
title: "Data Overview"
author: "Philipp Sven Lars Schäfer"
date: "`r format(Sys.time(), '%d %B, %Y')`"
editor: source
engine: knitr
---

# Packages

```{r}
suppressPackageStartupMessages({
  library(tidyverse)
  library(ggdark)
  library(factoextra)
  library(FactoMineR)
  library(magick)
  library(ComplexHeatmap)
  library(circlize)
})

source(file.path("..", "src", "read_data.R"))
source(file.path("..", "src", "colors.R"))
```

# Data

```{r}
input_dir = file.path("..", "data")
```

```{r}
meta_df <- read_harmonized_meta_data(input_dir)
data_list <- read_harmonized_experimental_data(input_dir)
data_list <- filter_harmonized_experimental_data(meta_df, data_list)
celltype_meta <- read_celltype_meta(input_dir)
gene_meta <- read_gene_meta(input_dir)
```

# Checks

```{r}
# only works due to filtering step above
stopifnot(all(
  purrr::map_lgl(data_list, ~ all(.x$specimen_id %in% meta_df$specimen_id)))
)

# no subject is recorded in more than one dataset
(meta_df %>%
  dplyr::count(subject_id, dataset) %>%
  dplyr::count(subject_id) %>%
  dplyr::pull(n) == 1) %>%
  all() %>%
  stopifnot()

# we have the baseline specimen (planned_day_relative_to_boost = 0) for every subject
(meta_df %>%
  dplyr::select(subject_id, planned_day_relative_to_boost, dataset) %>%
  dplyr::distinct() %>%
  dplyr::mutate(is_baseline = (planned_day_relative_to_boost==0)) %>%
  dplyr::group_by(subject_id) %>%
  dplyr::summarize(baseline_present = any(is_baseline),
                   dataset = first(dataset)) %>%
  dplyr::group_by(dataset) %>%
  dplyr::summarize(baseline_present_frac = mean(baseline_present)) %>%
  pull(baseline_present_frac) == 1) %>%
  all() %>%
  stopifnot()
```

# Demography & Metadata

## How many subjects do we have per dataset / per partition?

```{r}
meta_df %>%
  dplyr::select(subject_id, dataset, partition, infancy_vac) %>%
  dplyr::distinct() %>%
  dplyr::count(dataset, partition, infancy_vac) %>%
  ggplot(aes(x = n, y = dataset, fill = infancy_vac, color = partition)) +
  geom_col(position = position_dodge(width = 0.9), width = 0.7, alpha=0.5) +  
  geom_text(aes(label = n), 
            position = position_dodge(width = 0.9),  
            hjust = -0.2, vjust = 0.5, size = 3, color = "white") +
  scale_fill_manual(values = infancy_vac_colors) +
  scale_color_manual(values = partition_colors) +
  ggdark::dark_mode(verbose = FALSE)
```

## What is the age range

```{r}
meta_df %>%
  ggplot() + 
  geom_histogram(aes(x=age_at_boost, fill=dataset, y=after_stat(density)), 
                 color="black", position="identity", bins=30) +
  facet_wrap(~dataset) +
  scale_fill_manual(values=dataset_colors) +
  ggdark::dark_mode(verbose=FALSE)
```

# Assays & Timepoints per Subject / Dataset

## What is the difference between planned and actual booster administration

```{r}
meta_df %>%
  ggplot() +
  geom_histogram(aes(x=diff_relative_to_boost), binwidth=1) +
  facet_wrap(~dataset) +
  ggdark::dark_mode(verbose=FALSE)
```

## What is the difference between planned and actual for the baseline

```{r}
meta_df %>%
  dplyr::filter(planned_day_relative_to_boost==0) %>%
  dplyr::mutate(`diff_>_15` = abs(actual_day_relative_to_boost) > 15) %>%
  ggplot() +
  geom_histogram(aes(x=actual_day_relative_to_boost, fill=`diff_>_15`), binwidth=1) +
  facet_wrap(~dataset) +
  ggdark::dark_mode(verbose=FALSE)
```

## How many time points do we have per subject?

```{r}
meta_df %>%
  dplyr::count(dataset, subject_id) %>%
  ggplot() +
  geom_histogram(aes(x=n, fill=dataset),color="black", binwidth=1) +
  facet_wrap(~dataset, ncol=2) +
  ggdark::dark_mode(verbose=FALSE) +
  scale_x_continuous(breaks=seq(0, 8, 1)) +
  scale_fill_manual(values=dataset_colors)
```

## How many assays do we have for the baseline measurement?

```{r}
assays_per_specimen <- purrr::imap(data_list, ~ .x %>% 
              dplyr::select(specimen_id) %>%
              dplyr::distinct() %>%
              dplyr::mutate(assay=.y)) %>%
  dplyr::bind_rows() %>%
  dplyr::left_join(meta_df, by="specimen_id") %>%
  dplyr::select(specimen_id, assay, subject_id, planned_day_relative_to_boost, infancy_vac, dataset)

assays_per_specimen %>%
  dplyr::count(subject_id, planned_day_relative_to_boost, dataset, specimen_id) %>%
  dplyr::filter(!is.na(subject_id)) %>%
  dplyr::filter(planned_day_relative_to_boost==0) %>%
  ggplot() +
  geom_histogram(aes(x=n, fill=dataset), color="black", binwidth=1) +
  facet_wrap(~dataset, ncol=2) +
  ggdark::dark_mode(verbose=FALSE) +
  scale_x_continuous(breaks=seq(0, 8, 1)) +
  scale_fill_manual(values=dataset_colors)
```

```{r fig.height=6, fig.width=12}
annotation_data <- meta_df %>%
  dplyr::select(specimen_id, planned_day_relative_to_boost, infancy_vac, dataset, biological_sex) %>%
  dplyr::filter(specimen_id %in% assays_per_specimen$specimen_id) %>%
  dplyr::filter(planned_day_relative_to_boost == 0) %>% 
  dplyr::select(- planned_day_relative_to_boost) %>%
  dplyr::arrange(dataset, infancy_vac, biological_sex) %>%
  tibble::column_to_rownames("specimen_id") %>%
  dplyr::select(dataset, infancy_vac, biological_sex)

heatmap_data <- assays_per_specimen %>%
  dplyr::select(specimen_id, assay) %>%
  dplyr::mutate(value = 1) %>%
  tidyr::pivot_wider(names_from="assay", values_from="value") %>%
  mutate(across(everything(), .fns = ~replace_na(.,0))) %>%
  tibble::column_to_rownames("specimen_id") %>%
  as.matrix() %>%
  t()

colnames(heatmap_data) <- as.character(colnames(heatmap_data))
heatmap_data <- heatmap_data[, rownames(annotation_data)]

# Create the heatmap
ht <- Heatmap(heatmap_data, 
              name = "heatmap", 
              row_title = "", 
              column_title = "Baseline (Planned Day 0) Specimens",
              col = colorRamp2(c(0, 1), c("white", "black")),
              cluster_rows = FALSE,
              cluster_columns = FALSE,
              show_row_names = TRUE, 
              show_column_names = FALSE,
              width = unit(16, "cm"),
              top_annotation = ComplexHeatmap::HeatmapAnnotation(df = annotation_data,
                                    col = list(
                                      "dataset" = dataset_colors,
                                      "infancy_vac" = infancy_vac_colors,
                                      "biological_sex" = sex_colors
                                    ),
                                    which="column")
)

draw(ht, 
     heatmap_legend_side = "top",
     annotation_legend_side = "top",
     show_heatmap_legend = FALSE # don't show colorbar
     )
```

# Training Data

## How many subjects can we use to construct training data?

### 1) Antibody Level Tasks (anti-PT 14 days after booster)

- All but 7 people were assayed approximately 14 days after the booster administration

```{r}
meta_df %>%
  dplyr::filter(planned_day_relative_to_boost==14) %>%
  ggplot(aes(x=diff_relative_to_boost)) +
  geom_histogram(binwidth=1, color = "white", fill = "blue", alpha=0.1) +
  stat_bin(aes(label = after_stat(count)), binwidth=1, geom = "text", vjust = -0.5, color = "white", size = 3) +  
  ggdark::dark_mode(verbose=FALSE) +
  labs(x="Fraction of NA Values per Specimen")
```

```{r}
meta_df %>%
  dplyr::select(subject_id, infancy_vac, dataset, biological_sex, ethnicity, race, year_of_birth) %>%
  dplyr::distinct() %>%
  dplyr::mutate(task_1_possible = 
                  subject_id %in% (meta_df %>%
                                     dplyr::filter(planned_day_relative_to_boost==14) %>%
                                     dplyr::left_join(data_list$plasma_antibody_levels, by="specimen_id") %>%
                                     dplyr::filter(isotype_antigen=="IgG_PT") %>%
                                     tidyr::drop_na() %>%
                                     dplyr::pull(subject_id))
  ) %>%
  dplyr::group_by(dataset) %>%
  dplyr::summarise(task_1_possible_total = sum(task_1_possible), 
                   task_1_possible_fraction = mean(task_1_possible))
```

### 2) Cell Frequency Tasks (Monocytes 1 day after booster)

- But there is one person where it does not really make sense

```{r}
meta_df %>%
  dplyr::filter(planned_day_relative_to_boost==1) %>%
  ggplot(aes(x=diff_relative_to_boost)) +
  geom_histogram(binwidth=1, color = "white", fill = "blue", alpha=0.1) +
  stat_bin(aes(label = after_stat(count)), binwidth=1, geom = "text", vjust = -0.5, color = "white", size = 3) +  
  ggdark::dark_mode(verbose=FALSE) +
  labs(x="Fraction of NA Values per Specimen")
```

```{r}
meta_df %>%
  dplyr::select(subject_id, infancy_vac, dataset, biological_sex, ethnicity, race, year_of_birth) %>%
  dplyr::distinct() %>%
  dplyr::mutate(task_possible = 
                  subject_id %in% (meta_df %>%
                                     dplyr::filter(planned_day_relative_to_boost==1) %>%
                                     dplyr::left_join(data_list$pbmc_cell_frequency, by="specimen_id") %>%
                                     dplyr::filter(cell_type_name=="Monocytes") %>%
                                     tidyr::drop_na() %>%
                                     dplyr::pull(subject_id))
  ) %>%
  dplyr::group_by(dataset) %>%
  dplyr::summarise(task_possible_total = sum(task_possible), 
                   task_possible_fraction = mean(task_possible))
```

### 3) Gene Expression Tasks (CCL3 expression 3 days after booster)

- There are 7 people for which this task does nto make sense

```{r}
meta_df %>%
  dplyr::filter(planned_day_relative_to_boost==3) %>%
  ggplot(aes(x=diff_relative_to_boost)) +
  geom_histogram(binwidth=1, color = "white", fill = "blue", alpha=0.1) +
  stat_bin(aes(label = after_stat(count)), binwidth=1, geom = "text", vjust = -0.5, color = "white", size = 3) +  
  ggdark::dark_mode(verbose=FALSE) +
  labs(x="Fraction of NA Values per Specimen")
```

```{r}
all_genes <- unique(data_list$pbmc_gene_expression$versioned_ensembl_gene_id)
ccl3_ensembl <- "ENSG00000277632"
ccl3_ensembl_versioned <- all_genes[str_starts(all_genes, ccl3_ensembl)]

meta_df %>%
  dplyr::select(subject_id, infancy_vac, dataset, biological_sex, ethnicity, race, year_of_birth) %>%
  dplyr::distinct() %>%
  dplyr::mutate(task_possible = 
                  subject_id %in% (meta_df %>%
                                     dplyr::filter(planned_day_relative_to_boost==3) %>%
                                     dplyr::left_join(data_list$pbmc_gene_expression, by="specimen_id") %>%
                                     dplyr::filter(versioned_ensembl_gene_id==ccl3_ensembl_versioned) %>%
                                     tidyr::drop_na() %>%
                                     dplyr::pull(subject_id))
  ) %>%
  dplyr::group_by(dataset) %>%
  dplyr::summarise(task_possible_total = sum(task_possible), 
                   task_possible_fraction = mean(task_possible))
```

# Missing Data

## How many NA values do we have per assay?

```{r}
value_vars_per_modality <- list(
  "pbmc_cell_frequency" = c("percent_live_cell"),
  "pbmc_gene_expression" = c("raw_count", "tpm"),
  "plasma_antibody_levels" = c("MFI", "MFI_normalised"),
  "plasma_cytokine_concentrations_by_legendplex" = c("concentration"),
  "plasma_cytokine_concentrations_by_olink" = c("concentration"),
  "t_cell_activation" = c("analyte_percentages"),
  "t_cell_polarization" = c("analyte_counts")
)

purrr::imap_dfr(data_list, function(df, modality) {
  purrr::map_dfr(value_vars_per_modality[[modality]], 
                 ~ tibble(modality = modality,
                          value_vars = .x,
                          sum_na = sum(is.na(df[[.x]]))))
})
```

## What is the fraction of NA values in the PBMC Cell Type Frequencies

### Per Cell Type

```{r}
data_list$pbmc_cell_frequency %>%
  dplyr::mutate(is_na = is.na(percent_live_cell)) %>%
  dplyr::group_by(cell_type_name) %>%
  dplyr::summarise(frac_na = sum(is_na) / n()) %>%
  dplyr::arrange(desc(frac_na)) %>%
  dplyr::mutate(cell_type_name = factor(cell_type_name, levels=rev(cell_type_name))) %>%
  ggplot() +
  geom_col(aes(y=cell_type_name, x=frac_na)) +
  ggdark::dark_mode(verbose=FALSE) +
  labs(x="Fraction of NA Values per Cell Type")
```

### Per Cell Specimen

```{r}
data_list$pbmc_cell_frequency %>%
  dplyr::mutate(is_na = is.na(percent_live_cell)) %>%
  dplyr::group_by(specimen_id) %>%
  dplyr::summarise(frac_na = sum(is_na) / n()) %>%
  ggplot(aes(x = frac_na)) +
  geom_histogram(bins = 20, color = "white", fill = "blue", alpha=0.1) +  # Add color for better visibility
  stat_bin(aes(label = after_stat(count)), bins = 20, geom = "text", 
           vjust = -0.5, color = "white", size = 3) +  # Display bin counts at the top
  ggdark::dark_mode(verbose=FALSE) +
  labs(x="Fraction of NA Values per Specimen")
```

So there are 5 specimen with more than 20% NA, which are all from the same subject!

```{r}
data_list$pbmc_cell_frequency %>%
  dplyr::mutate(is_na = is.na(percent_live_cell)) %>%
  dplyr::group_by(specimen_id) %>%
  dplyr::summarise(frac_na = sum(is_na) / n()) %>%
  dplyr::filter(frac_na > 0.2) %>%
  dplyr::left_join(meta_df, by="specimen_id")
```

```{r fig.width=12, include=FALSE}
data_list$pbmc_cell_frequency %>%
  dplyr::filter(specimen_id %in% c(537, 538, 539, 540, 541)) %>%
  dplyr::left_join(meta_df, by="specimen_id") %>%
  tidyr::drop_na() %>%
  ggplot() +
  geom_line(aes(x=planned_day_relative_to_boost, y=percent_live_cell, color=cell_type_name)) +
  ggdark::dark_mode(verbose=FALSE)
```

## What is the fraction of NA values in the Olink Data?

### Per Protein

```{r fig.height=6}
data_list$plasma_cytokine_concentrations_by_olink %>%
  dplyr::mutate(is_na = is.na(concentration)) %>%
  dplyr::group_by(protein_id) %>%
  dplyr::summarise(frac_na = sum(is_na) / n()) %>%
  dplyr::arrange(desc(frac_na)) %>%
  dplyr::mutate(protein_id = factor(protein_id, levels=rev(protein_id))) %>%
  ggplot() +
  geom_col(aes(y=protein_id, x=frac_na)) +
  ggdark::dark_mode(verbose=FALSE) +
  labs(x="Fraction of NA Values per Protein")
```

### Per Specimen

```{r}
data_list$plasma_cytokine_concentrations_by_olink %>%
  dplyr::mutate(is_na = is.na(concentration)) %>%
  dplyr::group_by(specimen_id) %>%
  dplyr::summarise(frac_na = sum(is_na) / n()) %>%
  ggplot(aes(x = frac_na)) +
  geom_histogram(bins = 20, color = "white", fill = "blue", alpha=0.1) +  # Add color for better visibility
  stat_bin(aes(label = after_stat(count)), bins = 20, geom = "text", 
           vjust = -0.5, color = "white", size = 3) +  # Display bin counts at the top
  ggdark::dark_mode(verbose=FALSE) +
  labs(x="Fraction of NA Values per Specimen")
```

```{r}
data_list$plasma_cytokine_concentrations_by_olink %>%
  dplyr::mutate(is_na = is.na(concentration)) %>%
  dplyr::group_by(specimen_id) %>%
  dplyr::summarise(frac_na = sum(is_na) / n()) %>%
  dplyr::filter(frac_na > 0.2) %>%
  dplyr::left_join(meta_df, by="specimen_id") %>%
  dplyr::select(specimen_id, frac_na, subject_id, dataset, 
                actual_day_relative_to_boost, planned_day_relative_to_boost)
```

# Question for Assay Metadata

## How does the quality control for Olink look like?

```{r}
data_list$plasma_cytokine_concentrations_by_olink %>%
  dplyr::count(quality_control)
```


```{r}
data_list$plasma_cytokine_concentrations_by_olink %>%
  dplyr::left_join(meta_df, by="specimen_id") %>%
  dplyr::count(partition, dataset)

data_list$plasma_cytokine_concentrations_by_legendplex %>%
  dplyr::left_join(meta_df, by="specimen_id") %>%
  dplyr::count(partition, dataset)
```

## How often is the Cytokine Concentration below the LOD in Olink

```{r}
# check how often is it below the limit of detection?
data_list$plasma_cytokine_concentrations_by_olink %>%
  dplyr::mutate(below_lod = concentration < lower_limit_of_quantitation) %>%
  dplyr::count(below_lod)
```

## What is the fraction of unspecific antibody measurements

### Per Isotype-Antigen

```{r}
data_list$plasma_antibody_levels %>%
  dplyr::group_by(isotype_antigen) %>%
  dplyr::summarize(is_antigen_specific_fraction = mean(is_antigen_specific)) %>%
  ggplot(aes(x=is_antigen_specific_fraction)) +
  geom_histogram(bins=30, color = "white", fill = "blue", alpha=0.1) +
  stat_bin(aes(label = after_stat(count)), bins=30, geom = "text", vjust = -0.5, color = "white", size = 3) +  
  ggdark::dark_mode(verbose=FALSE)
```

### Per Specimen

So in some specimen we have non-specific antibody measurements? What does that mean?

```{r}
data_list$plasma_antibody_levels %>%
  dplyr::group_by(specimen_id) %>%
  dplyr::summarize(is_antigen_specific_fraction = mean(is_antigen_specific)) %>%
  ggplot(aes(x=is_antigen_specific_fraction)) +
  geom_histogram(binwidth=1, color = "white", fill = "blue", alpha=0.1) +
  stat_bin(aes(label = after_stat(count)), binwidth=1, geom = "text", vjust = -0.5, color = "white", size = 3) +  
  ggdark::dark_mode(verbose=FALSE)
```

# Hierarchy in PBMC Frequency Data

## NK

There are 3 types of NK cells:

- NK cells (CD3-CD19-CD56+): CD19-CD3-CD56+/++
- NK: CD19-CD3-CD56+
- CD56high NK cells: CD19-CD3-CD56++

And I don't know what to do with that information!

## No Gating Info for 2023

- I just assume it is the same as in 2022, but this might not be true since the data look so different

## No Gating Info for Basophils and CD19 in 2020

- I just assume it is the same as in 2021 and 2022

## No Gating Info for non-pDCs at all 

Not sure how to fix this)

```{r}
data_list$pbmc_cell_frequency %>%
  dplyr::left_join((meta_df %>%
                      dplyr::select(specimen_id, dataset) %>%
                      dplyr::distinct()),
                   by="specimen_id") %>%
  dplyr::left_join((celltype_meta %>%
                      dplyr::select(cell_type_name, dataset, level) %>%
                      dplyr::distinct()),
                    by=c("dataset", "cell_type_name")) %>%
  dplyr::filter(is.na(level)) %>%
  dplyr::select(dataset, cell_type_name) %>%
  dplyr::distinct()
```

## Should the Proportions sum up to 1 at level 0?

- Trying to take the hierarchical information into account, it still does not make sense

```{r}
data_list$pbmc_cell_frequency %>%
  dplyr::left_join((meta_df %>%
                      dplyr::select(specimen_id, dataset) %>%
                      dplyr::distinct()),
                   by="specimen_id") %>%
  dplyr::left_join((celltype_meta %>%
                      dplyr::select(cell_type_name, dataset, level) %>%
                      dplyr::distinct()),
                    by=c("dataset", "cell_type_name")) %>%
  dplyr::filter(level == 0) %>%
  dplyr::group_by(specimen_id) %>%
  dplyr::summarize(percent_live_cell_sum = sum(percent_live_cell, na.rm=TRUE)) %>%
  dplyr::left_join(meta_df, by="specimen_id") %>%
  ggplot() +
  geom_violin(aes(x=dataset, y=percent_live_cell_sum)) +
  ggdark::dark_mode(verbose=FALSE)
```

## Difference in 2023 Dataset

- It seems like the 2023 data are very very different, which makes me hesitant to use it for any kind of predictions on the challenge data. E.g. just looking at level 0 which should make sense I get the following

```{r}
# first check for NAs for the levels
data_list$pbmc_cell_frequency %>%
  dplyr::left_join((meta_df %>%
                      dplyr::select(specimen_id, dataset) %>%
                      dplyr::distinct()),
                   by="specimen_id") %>%
  dplyr::left_join((celltype_meta %>%
                      dplyr::select(cell_type_name, dataset, level) %>%
                      dplyr::distinct()),
                    by=c("dataset", "cell_type_name")) %>%
  dplyr::filter(is.na(level))
```

```{r}
data_list$pbmc_cell_frequency %>%
  dplyr::left_join((meta_df %>%
                      dplyr::select(specimen_id, dataset) %>%
                      dplyr::distinct()),
                   by="specimen_id") %>%
  dplyr::left_join((celltype_meta %>%
                      dplyr::select(cell_type_name, dataset, level) %>%
                      dplyr::distinct()),
                    by=c("dataset", "cell_type_name")) %>%
  dplyr::filter(level == 0) %>%
  dplyr::group_by(specimen_id) %>%
  dplyr::summarize(percent_live_cell_sum = sum(percent_live_cell, na.rm=TRUE)) %>%
  dplyr::left_join(meta_df, by="specimen_id") %>%
  ggplot() +
  geom_violin(aes(x=dataset, y=percent_live_cell_sum)) +
  ggdark::dark_mode(verbose=FALSE)
```

- Especially, when completely ignoring the hierarchy in the gating information, I would expect to have more than 100% in `percelt_live_cell_sum`, but this is not true for the specimens from the 2023 dataset

```{r}
data_list$pbmc_cell_frequency %>%
  dplyr::left_join((meta_df %>%
                      dplyr::select(specimen_id, dataset) %>%
                      dplyr::distinct()),
                   by="specimen_id") %>%
  dplyr::left_join((celltype_meta %>%
                      dplyr::select(cell_type_name, dataset, level) %>%
                      dplyr::distinct()),
                    by=c("dataset", "cell_type_name")) %>%
  dplyr::group_by(specimen_id) %>%
  dplyr::summarize(percent_live_cell_sum = sum(percent_live_cell, na.rm=TRUE)) %>%
  dplyr::left_join(meta_df, by="specimen_id") %>%
  ggplot() +
  geom_violin(aes(x=dataset, y=percent_live_cell_sum)) +
  ggdark::dark_mode(verbose=FALSE)
```

- Or what I use a list of predefined cell types of interest?

# Other Notes

- Only from 2022 onwards do we have time-course baseline measurements, i.e. `-30, -14/15, 0`

```{r}
meta_df %>%
  dplyr::count(partition, dataset, planned_day_relative_to_boost) %>%
  dplyr::filter(planned_day_relative_to_boost < 0)
```

# Appendix

<button class="accordion-button" type="button" data-bs-toggle="collapse" data-bs-target="#collapseOne" >Session Information</button><div id="collapseOne" class="accordion-collapse collapse"><div>

```{r}
sessionInfo()
```

</div></div>