---
title: "Model Selection"
author: "Philipp Sven Lars Schäfer"
date: "`r format(Sys.time(), '%d %B, %Y')`"
editor: source
engine: knitr
---

# Packages

```{r}
suppressPackageStartupMessages({
  library(tidyverse)
  library(flextable)
  library(ggdark)
  library(magick)
  library(glmnet)
  library(ranger)
  source(file.path("..", "src", "read_data.R"))
  source(file.path("..", "src", "colors.R"))
  source(file.path("..", "src", "generate_targets.R"))
  source(file.path("..", "src", "model.R"))
  source(file.path("..", "src", "normalize_integrate.R"))
})
```

# Data

```{r}
input_dir = file.path("..", "data")
output_dir = file.path("..", "results", "model_selection_no2020")
dir.create(output_dir, showWarnings = FALSE, recursive = TRUE)
```

```{r}
celltype_meta <- read_celltype_meta(input_dir)
gene_meta <- read_gene_meta_plus(input_dir)
protein_meta <- read_protein_meta(input_dir)

meta_data <- read_harmonized_meta_data(input_dir)
specimen_per_day <- get_specimen_per_day(meta_data=meta_data)

RECOMPUTE <- TRUE
if (RECOMPUTE) {
  raw_experimental_data <- read_raw_experimental_data(input_dir)
  
  filtered_experimental_data <- filter_experimental_data(
    meta_data=meta_data, 
    experimental_data=raw_experimental_data,
    gene_meta=gene_meta)
  
  write_rds(filtered_experimental_data, 
            file = file.path(input_dir, "prc_datasets", 
                             "filtered_experimental_data.RDS"))
  
  specimen_to_assay <- purrr::map(list("raw"=raw_experimental_data, "filtered"=filtered_experimental_data), function(exp_data) {
    # exp_data <- filtered_experimental_data; exp_name = "filtered"
    purrr::imap(exp_data, function(exp_df, exp_name) {
      exp_df %>%
        dplyr::select(specimen_id) %>%
        dplyr::distinct() %>%
        dplyr::mutate(assay = exp_name, specimen_id = as.numeric(specimen_id))
    }) %>%
      dplyr::bind_rows() %>%
      dplyr::mutate(present = TRUE) %>%
      tidyr::pivot_wider(names_from=assay, values_from=present) %>%
      mutate_all(~ ifelse(is.na(.), FALSE, .))
  })
  
  meta_data_plus_assays <- purrr::map(specimen_to_assay, function(df) {
    meta_data %>%
      dplyr::left_join(df, by="specimen_id") %>%
      mutate_all(~ ifelse(is.na(.), FALSE, .))
  })
  
  write_rds(specimen_to_assay, 
            file = file.path(input_dir, "prc_datasets", 
                             "specimen_to_assay.RDS"))
  write_rds(meta_data_plus_assays, 
            file = file.path(input_dir, "prc_datasets", 
                             "meta_data_plus_assays.RDS"))
  
  normalized_experimental_data <- normalize_experimental_data(
    meta_data=meta_data, 
    raw_experimental_data=filtered_experimental_data,
    gene_meta=gene_meta)
  
  write_rds(normalized_experimental_data, 
            file = file.path(input_dir, "prc_datasets", 
                             "normalized_experimental_data.RDS"))
  
  integrated_experimental_data <- integrate_experimental_data(
    meta_data=meta_data, 
    normalized_experimental_data=normalized_experimental_data)
  
  write_rds(integrated_experimental_data, 
            file = file.path(input_dir, "prc_datasets", 
                             "integrated_experimental_data.RDS"))

  # use raw/filtered experimental data for computation of targets
  target_list <- generate_all_targets(
    meta_data=meta_data, 
    experimental_data=filtered_experimental_data, 
    experimental_data_settings=experimental_data_settings, 
    gene_meta=gene_meta,
    protein_meta=protein_meta
    )
  
  # use raw/filtered experimental data for computation of targets
  baseline_list <- generate_all_baselines(
    meta_data=meta_data, 
    experimental_data=filtered_experimental_data, 
    experimental_data_settings=experimental_data_settings, 
    gene_meta=gene_meta,
    protein_meta=protein_meta
    )
  
  stopifnot(identical(names(target_list), names(baseline_list)))
  purrr::map_lgl(names(target_list), function(task_name) {
    (target_list[[task_name]] %>%
      dplyr::inner_join(baseline_list[[task_name]], by="subject_id") %>%
      dplyr::summarise(cor = cor(baseline.x, baseline.y)) %>%
      dplyr::pull(cor)) == 1
  }) %>%
    all() %>%
    stopifnot()
  
  write_rds(target_list, file = file.path(input_dir, "prc_datasets", 
                                          "target_list.RDS"))
  
  write_rds(baseline_list, file = file.path(input_dir, "prc_datasets", 
                                            "baseline_list.RDS"))
  
  rm(raw_experimental_data, filtered_experimental_data)
  experimental_data <- integrated_experimental_data
  experimental_data <- experimental_data[-which(names(experimental_data) ==
                                                "pbmc_gene_expression_counts")]
} else {
  experimental_data <- read_rds(file = file.path(input_dir, "prc_datasets", 
                                                            "integrated_experimental_data.RDS"))
  experimental_data <- experimental_data[-which(names(experimental_data) ==
                                                "pbmc_gene_expression_counts")]
  
  target_list <- read_rds(file = file.path(input_dir, "prc_datasets",  "target_list.RDS"))
  
  baseline_list <- read_rds(file = file.path(input_dir, "prc_datasets", "baseline_list.RDS"))
  
  specimen_to_assay <- read_rds(file = file.path(input_dir, "prc_datasets", "specimen_to_assay.RDS"))
  
  meta_data_plus_assays <- read_rds(file = file.path(input_dir, "prc_datasets", "meta_data_plus_assays.RDS"))
}
```

```{r}
specimen_2020 <- meta_data %>%
  dplyr::filter(dataset == "2020_dataset") %>%
  dplyr::pull(specimen_id) %>%
  unique()

subjects_2020 <- meta_data %>%
  dplyr::filter(dataset == "2020_dataset") %>%
  dplyr::pull(subject_id) %>%
  unique()

experimental_data <- purrr::map(experimental_data, ~ .x %>% dplyr::filter(!(as.numeric(specimen_id) %in% specimen_2020)))

target_list <- purrr::map(target_list, ~ .x %>% dplyr::filter(!(subject_id %in% subjects_2020)))

baseline_list <- purrr::map(baseline_list, ~ .x %>% dplyr::filter(!(subject_id %in% subjects_2020)))

specimen_to_assay <- purrr::map(specimen_to_assay, ~ .x %>% dplyr::filter(!(specimen_id %in% specimen_2020)))

meta_data_plus_assays <- purrr::map(meta_data_plus_assays, ~ .x %>% dplyr::filter(!(subject_id %in% subjects_2020)))
```

# Tasks

```{r}
task_meta <- list(
  task_11 = list(
    name = "task_11",
    header = "## Task 1.1",
    description = "Rank the individuals by IgG antibody levels against pertussis toxin (PT) that we detect in plasma 14 days post booster vaccinations."
  ),
  task_12 = list(
    name = "task_12",
    header = "## Task 1.2",
    description = "Rank the individuals by fold change of IgG antibody levels against pertussis toxin (PT) that we detect in plasma 14 days post booster vaccinations compared to titer values at day 0."
  ),
  task_21 = list(
    name = "task_21",
    header = "## Task 2.1",
    description = "Rank the individuals by predicted frequency of Monocytes on day 1 post boost after vaccination."
  ),
  task_22 = list(
    name = "task_22",
    header = "## Task 2.2",
    description = "Rank the individuals by fold change of predicted frequency of Monocytes on day 1 post booster vaccination compared to cell frequency values at day 0."
  ),
  task_31 = list(
    name = "task_31",
    header = "## Task 3.1",
    description = "Rank the individuals by predicted gene expression of CCL3 on day 3 post-booster vaccination."
  ),
  task_32 = list(
    name = "task_32",
    header = "## Task 3.2",
    description = "Rank the individuals by fold change of predicted gene expression of CCL3 on day 3 post booster vaccination compared to gene expression values at day 0."
  ),
  task_41 = list(
    name = "task_41",
    header = "## Task 4.1",
    description = "Rank the individuals based on their Th1/Th2 (IFN-g/IL-5) polarization ratio on Day 30 post-booster vaccination."
  )
)
```

# Rationale

- How to choose the right model?

  - Top ranking performance
  
  - Low variance between the test sets
  
  - If several models with top performance and low variance, choose the more regularized model
  
  - If several models with top performance and low variance, choose model that is not using assay that is missing often in test data (-> Olink!)

# Conclusions

TODO

# Prepare Features

```{r}
# get the 2000 most variable genes based on the mean dispersion (variance/mean) across cohorts
N_HVG <- 2000
hvg <- gene_meta %>% 
  dplyr::mutate("prefix_versioned_ensembl_gene_id_clean" =
                  paste0("pbmc_gene_expression_", versioned_ensembl_gene_id_clean)) %>%
  dplyr::slice_max(mean_rank, n=N_HVG, with_ties=TRUE)

meta_data_cov <- get_metadata_covariates(meta_data) %>%
  dplyr::select(-c(ethnicity_Not_Hispanic_or_Latino, ethnicity_Hispanic_or_Latino))

experimental_predictors_day_0 <- 
  generate_wide_experimental_data(experimental_data=experimental_data,
                                  impute="knn", 
                                  max_NA_frac_sample = 0.5,
                                  max_NA_frac_feature = 0.25,
                                  verbose=TRUE) %>%
  purrr::imap(function(df, modality) {
    colnames(df) <- paste0(modality, "_", colnames(df)) # since feature names can be repeated across assays
    df <- as.data.frame(df) %>%
      tibble::rownames_to_column("specimen_id") %>%
      dplyr::mutate(specimen_id = as.numeric(specimen_id)) %>%
      dplyr::inner_join((specimen_per_day$day_0 %>% 
                           dplyr::select(subject_id, specimen_id, dataset)),
                        by="specimen_id") %>%
      dplyr::select(-c(specimen_id, dataset)) %>%
      dplyr::select(subject_id, everything())
    return(df)
  })

experimental_predictors_day_0$pbmc_gene_expression <-
  experimental_predictors_day_0$pbmc_gene_expression[, c("subject_id", hvg$prefix_versioned_ensembl_gene_id_clean)]

#purrr::map(experimental_predictors_day_0, ~ dim(.x))

experimental_predictors_day_0_pca <- experimental_predictors_day_0 %>%
  purrr::imap(function(df, modality) {
    # df <- experimental_predictors_day_0$
    mtx <- df %>% dplyr::select(-subject_id) %>% as.matrix()
    pca <- stats::prcomp(mtx)
    sdev_ratio <- cumsum(pca$sdev) / sum(pca$sdev)
    n_pcs <- max(which(sdev_ratio < 0.9))
    df_pca <- as.data.frame(pca$x[, 1:n_pcs]) %>%
      dplyr::mutate(subject_id = df$subject_id) %>%
      dplyr::select(subject_id, everything())
    return(df_pca)
})

modality_powerset <- purrr::map(1:length(experimental_predictors_day_0), function(i) {
  utils::combn(names(experimental_predictors_day_0), i, simplify = FALSE)
}) %>%
   unlist(recursive = FALSE)

#purrr::map(experimental_predictors_day_0_pca, ~ dim(.x))

meta_data_plus_assays_day_0 <- meta_data_plus_assays$filtered %>%
  dplyr::filter(specimen_id %in% specimen_per_day$day_0$specimen_id)
stopifnot(length(meta_data_plus_assays_day_0$subject_id) == 
            length(unique(meta_data_plus_assays_day_0$subject_id)))
```

Check which assays are missing the least often?

- Olink is missing quite often, because I removed all measurements for the 2020 cohort

```{r}
# which assays are most/least often missing?
meta_data_plus_assays_day_0 %>%
  dplyr::select(dplyr::all_of(names(experimental_predictors_day_0))) %>%
  dplyr::mutate(row_sum = apply(select(., matches(".+")), 1, sum)) %>%
  dplyr::filter(row_sum > 0) %>%
  tidyr::pivot_longer(cols=!row_sum) %>%
  dplyr::group_by(name) %>%
  dplyr::summarise(mean = mean(value)) %>%
  dplyr::arrange(desc(mean))
```

```{r}
MAX_N_ASSAYS_MISSING <- 2
K_FOR_KNN_IMPUTE <- 7

feature_list <- list(
  "metadata" = meta_data_cov,
  "metadata+baseline" = meta_data_cov # will take care of that logic in loop
)

# check for each feature setting (aka modality setting)
# whether to many assays are missing for any subject 
# (by default we remove subject if more than 2 of the required feature sets are missing)
# usually we exclude here subject from the 2020 dataset
for (modality_set in modality_powerset) {
  # modality_set <- modality_powerset[[100]]
  modality_set_name <- paste0(paste0(modality_set, collapse = "_pca+"), "_pca+metadata+baseline")
  
  # inner join here, other subject have no assays for day 0...
  modality_set_check_assays  <- meta_data_cov %>%
    dplyr::inner_join(dplyr::select(meta_data_plus_assays_day_0, dplyr::all_of(c("subject_id", modality_set))), 
                        by="subject_id")
  modality_set_check_assays[, modality_set][is.na(modality_set_check_assays[, modality_set])] <- FALSE
  modality_set_check_assays[, "n_assays_present"] <- rowSums(modality_set_check_assays[, modality_set])
  modality_set_check_assays[, "n_assays_missing"] <- length(modality_set) - modality_set_check_assays[, "n_assays_present"]
  #modality_set_check_assays[, "n_assays_max"] <- length(modality_set)
  #modality_set_check_assays <- modality_set_check_assays %>% dplyr::mutate(n_assays_frac = n_assays_present / n_assays_max)
  
  subjects_to_exclude <- modality_set_check_assays %>%
    dplyr::filter(n_assays_missing >= !!MAX_N_ASSAYS_MISSING) %>%
    dplyr::pull(subject_id)
  
  modality_set_meta_df <- meta_data_cov %>%
    dplyr::filter(!(subject_id %in% subjects_to_exclude))
  
  modality_set_df <- 
    experimental_predictors_day_0_pca[modality_set] %>%
    purrr::reduce(full_join, by="subject_id") %>%
    dplyr::inner_join(modality_set_meta_df, by="subject_id")
  
  modality_set_mtx <- modality_set_df %>%
    tibble::column_to_rownames("subject_id") %>%
    as.matrix() %>%
    t() %>%
    impute::impute.knn(data=., k=K_FOR_KNN_IMPUTE, maxp=10000, colmax=1) %>%
    .$data %>%
    t()
  
  feature_list[[modality_set_name]] <- as.data.frame(modality_set_mtx) %>%
    tibble::rownames_to_column("subject_id") %>%
    dplyr::mutate(subject_id = as.numeric(subject_id))
}

# adding power set of single-omic predictors
for (modality_set in modality_powerset) {
  # modality_set <- modality_powerset[[100]]
  modality_set_name <- paste0(paste0(modality_set, collapse = "+"), "+metadata+baseline")
  
# inner join here, other subject have no assays for day 0...
  modality_set_check_assays  <- meta_data_cov %>%
    dplyr::inner_join(dplyr::select(meta_data_plus_assays_day_0, dplyr::all_of(c("subject_id", modality_set))), 
                        by="subject_id")
  modality_set_check_assays[, modality_set][is.na(modality_set_check_assays[, modality_set])] <- FALSE
  modality_set_check_assays[, "n_assays_present"] <- rowSums(modality_set_check_assays[, modality_set])
  modality_set_check_assays[, "n_assays_missing"] <- length(modality_set) - modality_set_check_assays[, "n_assays_present"]
  #modality_set_check_assays[, "n_assays_max"] <- length(modality_set)
  #modality_set_check_assays <- modality_set_check_assays %>% dplyr::mutate(n_assays_frac = n_assays_present / n_assays_max)
  
  
  subjects_to_exclude <- modality_set_check_assays %>%
    dplyr::filter(n_assays_missing >= !!MAX_N_ASSAYS_MISSING) %>%
    dplyr::pull(subject_id)
  
  modality_set_meta_df <- meta_data_cov %>%
    dplyr::filter(!(subject_id %in% subjects_to_exclude))
  
  modality_set_df <- 
    experimental_predictors_day_0[modality_set] %>%
    purrr::reduce(full_join, by="subject_id") %>%
    dplyr::inner_join(modality_set_meta_df, by="subject_id")
  
  modality_set_mtx <- modality_set_df %>%
    tibble::column_to_rownames("subject_id") %>%
    as.matrix() %>%
    t() %>%
    impute::impute.knn(data=., k=K_FOR_KNN_IMPUTE, maxp=10000, colmax=1) %>%
    .$data %>%
    t()
  
  feature_list[[modality_set_name]] <- as.data.frame(modality_set_mtx) %>%
    tibble::rownames_to_column("subject_id") %>%
    dplyr::mutate(subject_id = as.numeric(subject_id))
}

stopifnot(all(purrr::map_dbl(feature_list, ~ mean(is.na(.x))) == 0))

#purrr::map(feature_list, ~ dim(.x))
```

# Model Evaluation / Assessment

```{r}
set.seed(42)

tictoc::tic()

result_list <-  list()

for (task in task_meta) {
  # task <- task_meta[[5]]
  
  task_df <- target_list[[task$name]]
  
  target_df <- task_df %>%
    dplyr::select(subject_id, target)
  
  baseline_df <- task_df %>% 
    dplyr::select(subject_id, baseline)
  
  baseline_model_df <- task_df %>%
    dplyr::left_join(dplyr::distinct(dplyr::select(meta_data, c(subject_id, dataset))),
                     by="subject_id")
  
  all_datasets <- unique(baseline_model_df$dataset)
  
  for (test_dataset in all_datasets) {
    result_list <- rlist::list.append(result_list, tibble::tibble(
      task = task$name,
      trainset = NA,
      testset = test_dataset,
      test_subset = task_df$subject_id[baseline_model_df$dataset==test_dataset],
      target = task_df$target[baseline_model_df$dataset==test_dataset], 
      prediction = task_df$baseline[baseline_model_df$dataset==test_dataset],
      testset_mean = NA,
      model = "baseline",
      feature_set = "baseline"
    ))
  }
  # for (model_name in c("lasso", "elnet", "rf", "rf+boruta")) {
  for (model_name in c("lasso", "elnet", "rf")) {
    # model_name <- "lasso"
    for (feature_name in names(feature_list)) {
      # feature_name <- names(feature_list)[99]
      feature_df <- feature_list[[feature_name]]
      
      if (str_detect(feature_name, "baseline")) {
        feature_df <- feature_df %>%
          dplyr::left_join(baseline_df, by="subject_id")
      }
      
      model_df <- target_df %>%
        dplyr::left_join(feature_df, by="subject_id") %>%
        dplyr::left_join(dplyr::distinct(dplyr::select(meta_data, c(subject_id, dataset))),
                         by="subject_id")
      
      #stopifnot(!any(is.na(model_df))) # guard rail
      model_df <- model_df %>% tidyr::drop_na()
      stopifnot(nrow(model_df)==nrow(dplyr::distinct(model_df)))
      
      all_datasets <- unique(model_df$dataset)
      
      cross_dataset_out <- purrr::map(all_datasets, function(test_dataset) {
        # test_dataset <- all_datasets[2]
        train_dataset <- all_datasets[!all_datasets %in% test_dataset]
        
        train_df <- model_df %>%
          dplyr::filter(dataset %in% train_dataset) %>%
          dplyr::select(-c(dataset, subject_id))
        
        train_mtx <- as.matrix(dplyr::select(train_df, -target))
        train_target <- dplyr::pull(train_df, target)
        
        test_df_with_subject <- model_df %>%
          dplyr::filter(dataset %in% test_dataset)
        
        test_df <- test_df_with_subject %>%
          dplyr::select(-c(dataset, subject_id))
        
        test_mtx <- as.matrix(dplyr::select(test_df, -target))
        test_target <- dplyr::pull(test_df, target)
        
        if (model_name == "rf") {
          model <- ranger::ranger(formula = target ~ ., 
                                  data = train_df, 
                                  num.trees = 500)
          predictions <- predict(model, test_df)$predictions
        } 
        else if (model_name == "rf+boruta") {
          if (ncol(train_df) > 2) {
            boruta_out <- Boruta::Boruta(formula = target ~ .,
                                         data = train_df,
                                         maxRuns=1000,
                                         num.trees=500 
                                         )
            boruta_feats <- names(boruta_out$finalDecision)[boruta_out$finalDecision=="Confirmed"]
          } else {
            boruta_feats <- colnames(train_df)[colnames(train_df)!="target"]
          }
          if (length(boruta_feats) == 0) {
            predictions <- rep(mean(train_df$target), nrow(test_df))
          } else {
            train_df_subset <- train_df %>% dplyr::select(dplyr::all_of(c(boruta_feats, "target")))
            test_df_subset <- test_df %>% dplyr::select(dplyr::all_of(c(boruta_feats, "target")))
            model <- ranger::ranger(formula = target ~ ., 
                                    data = train_df_subset, 
                                    num.trees = 500)
            predictions <- predict(model, test_df_subset)$predictions
          }
        } 
        else if (model_name == "lasso") {
          # standardize the features, even though glmnet is doing this automatically!
          # feature_means <- matrixStats::colMeans2(train_mtx)
          # feature_sds <- matrixStats::colSds(train_mtx)
          # train_mtx <- t((t(train_mtx) - feature_means) / feature_sds)
          # train_target <- (train_target - mean(train_target)) / sd(train_target)
          #test_mtx <- t((t(test_mtx) - feature_means) / feature_sds)
          
          # inner hyperparameter optim loop using LOOCV to find best lambda
          model <- glmnet::cv.glmnet(x=train_mtx,
                                     y=train_target,
                                     nfolds=nrow(train_mtx),
                                     grouped=FALSE,
                                     alpha=1, # LASSO
                                     type.measure="mae", # not sure whether this is better
                                     standardize=TRUE, # by default each feature is standardized internally
                                     family="gaussian"
                                     )
          # investigate
          #beta_mtx <- as.data.frame(as.matrix(model$glmnet.fit$beta))
          #colnames(beta_mtx) <- paste0("lambda_", round(model$lambda, 3))
          #plot(model)
          #lambda_value <- model$lambda.1se, # model$lambda.min
          # never use zero features (with lamba=model$lambda[1]), since constant predictions
          # will have NA spearman...
          lambda_value <- ifelse(model$lambda.min==model$lambda[1], 
                                 model$lambda[2], model$lambda.min)
          predictions <- predict(model,
                                 newx=test_mtx, 
                                 s=lambda_value)[,1]
        }
        else if (model_name == "elnet") {
          # standardize the features, even though glmnet is doing this automatically!
          # feature_means <- matrixStats::colMeans2(train_mtx)
          # feature_sds <- matrixStats::colSds(train_mtx)
          # train_mtx <- t((t(train_mtx) - feature_means) / feature_sds)
          # train_target <- (train_target - mean(train_target)) / sd(train_target)
          #test_mtx <- t((t(test_mtx) - feature_means) / feature_sds)
          
          # inner hyperparameter optim loop using LOOCV to find best lambda
          model <- glmnet::cv.glmnet(x=train_mtx,
                                     y=train_target,
                                     nfolds=nrow(train_mtx),
                                     grouped=FALSE,
                                     alpha=0.5, # elastic net
                                     type.measure="mae", # not sure whether this is better
                                     standardize=TRUE, # by default each feature is standardized internally
                                     family="gaussian"
                                     )
          lambda_value <- ifelse(model$lambda.min==model$lambda[1], 
                                 model$lambda[2], model$lambda.min)
          predictions <- predict(model,
                                 newx=test_mtx, 
                                 s=lambda_value)[,1]
        }
        else {
          stop(paste0(model_name), " not implemented")
        }
        
        tibble::tibble(task=task$name,
                       trainset = paste0(train_dataset, collapse="__"),
                       testset = test_dataset,
                       test_subset = test_df_with_subject$subject_id,
                       target = test_df$target, 
                       prediction = predictions,
                       testset_mean = mean(test_df$target),
                       model = model_name,
                       feature_set = feature_name)
      }) %>%
        dplyr::bind_rows()
      result_list <- rlist::list.append(result_list, cross_dataset_out)
    }
  }
}

result_df <- dplyr::bind_rows(result_list)
result_df

tictoc::toc()
```

```{r}
result_summary <- result_df %>%
    dplyr::group_by(task, feature_set, model, trainset, testset) %>%
    dplyr::summarise(
      test_mse = get_mse(target, prediction),
      test_r2 = get_r2(target, prediction),
      test_srho = get_spearman(target, prediction),
      mse_same_cohort = get_mse(target, testset_mean),
      .groups = "drop"
    )
result_summary
```

Why is that? This happens when the boruta algorithm returns 0 important features, which means in the test set the predictions are set to some constant. And since a constant vector has 0 standard deviation, we cannot compute the correlation (where we divide by the standard deviation).

```{r}
result_summary %>%
  dplyr::filter(is.na(test_srho)) %>%
  dplyr::count(task, model)
```

```{r}
model_selection_df <- result_summary %>%
  dplyr::select(task, feature_set, model, testset, test_srho) %>%
  tidyr::pivot_wider(values_from=test_srho, names_from=testset) %>%
  dplyr::mutate(srho_mean = apply(select(., matches("[0-9]+_dataset")), 1, mean, na.rm=TRUE)) %>%
  dplyr::filter(!is.na(srho_mean)) %>% # so we only keep entries where predictions in all test sets where constant
  dplyr::mutate(srho_min = apply(select(., matches("[0-9]+_dataset")), 1, min, na.rm=TRUE)) %>%
  dplyr::mutate(srho_sd = apply(select(., matches("[0-9]+_dataset")), 1, sd, na.rm=TRUE)) %>%
  dplyr::mutate(srho_mean = round(srho_mean, 3)) %>%
  dplyr::mutate(srho_min = round(srho_min, 3)) %>%
  dplyr::mutate(srho_sd = round(srho_sd, 3)) %>%
  dplyr::mutate(across(matches("[0-9]+_dataset"), ~ round(.x, 3))) %>%
  dplyr::mutate(dim_red = dplyr::case_when(
    str_detect(feature_set, "_pca") ~ "pca",
    TRUE ~ "none"
  )) %>%
  dplyr::mutate(n_modalities = (stringr::str_count(feature_set, "\\+") + 1)) %>%
  dplyr::select(task, model, srho_mean, srho_sd, srho_min, !feature_set, feature_set)
```

```{r}
readr::write_delim(model_selection_df, file=file.path(output_dir, paste0(format(Sys.Date(), "%Y-%m-%d"), "_model_selection_df.tsv")), delim="\t")

#model_selection_df <- readr::read_delim(file.path(output_dir, "model_selection_df.tsv"), delim="\t")

library(openxlsx)
# see https://ycphs.github.io/openxlsx/reference/openxlsx_options.html
#base::options()
wb <- openxlsx::createWorkbook()
openxlsx::addWorksheet(wb, "results_unfiltered")
openxlsx::setColWidths(wb, sheet="results_unfiltered", cols = 1:10, widths = 15)
openxlsx::setColWidths(wb, sheet="results_unfiltered", cols = 11:12, widths = 100)
openxlsx::writeDataTable(wb, 
                         sheet = "results_unfiltered", 
                         x = model_selection_df, 
                         startRow = 1, 
                         startCol = 1,
                         tableStyle = "TableStyleDark1")

for (task in unique(model_selection_df$task)) {
  # task <- "task_21"
  task_result_df <- model_selection_df %>%
    dplyr::filter(task == !!task) %>%
    dplyr::filter(!is.na(srho_sd)) %>%
    dplyr::arrange(desc(srho_mean)) 
  openxlsx::addWorksheet(wb, task)
  openxlsx::setColWidths(wb, sheet=task, cols = 1:10, widths = 15)
  openxlsx::setColWidths(wb, sheet=task, cols = 11:12, widths = 100)
  openxlsx::writeDataTable(wb, 
                         sheet = task,
                         x = task_result_df, 
                         startRow = 1, 
                         startCol = 1,
                         tableStyle = "TableStyleDark1")
}
openxlsx::saveWorkbook(
  wb, 
  file=file.path(output_dir, paste0(format(Sys.Date(), "%Y-%m-%d"), "_model_selection_df.xlsx")), 
  overwrite = TRUE)
```

# Prediction for 2023

1. Load the submission template

```{r}
submission_df <- readr::read_tsv(file.path(input_dir, "3rdChallengeSubmissionTemplate_10032024.tsv"),
                                 show_col_types = FALSE)
submission_df
```

Check which assays are missing most often

```{r}
test_assays_available <- submission_df %>%
  dplyr::select(SubjectID) %>%
  tidylog::left_join(meta_data_plus_assays_day_0, by=c("SubjectID"="subject_id"))
```

```{r}
colMeans(test_assays_available[, names(experimental_predictors_day_0)]) %>% sort()
```

```{r}
rowSums(test_assays_available[, names(experimental_predictors_day_0)]) %>% sort()
```

2. Define for each task a list with the following entries

```{r}
colnames(submission_df)
```

```{r}
prediction_specs <- list(
  "task_11" = list(
    "model" = "baseline",
    "feature_set" = "baseline",
    "table_col" = "1.1) IgG-PT-D14-titer-Rank"
  ),
  "task_12" = list(
    "model" = "elnet",
    "feature_set" = "pbmc_cell_frequency+plasma_ab_titer+metadata+baseline",
    "table_col" = "1.2) IgG-PT-D14-FC-Rank"
  ),
  "task_21" = list(
    "model" = "rf",
    "feature_set" = "pbmc_cell_frequency_pca+plasma_cytokine_concentration_by_legendplex_pca+t_cell_activation_pca+metadata+baseline",
    "table_col" = "2.1) Monocytes-D1-Rank"
  ),
  "task_22" = list(
    "model" = "elnet",
    "feature_set" = "plasma_cytokine_concentration_by_legendplex+t_cell_activation+t_cell_polarization+metadata+baseline",
    "table_col" = "2.2) Monocytes-D1-FC-Rank"
  ),
  "task_31" = list(
    "model" = "lasso",
    "feature_set" = "plasma_cytokine_concentration_by_legendplex+metadata+baseline",
    "table_col" = "3.1) CCL3-D3-Rank"
  ),
  "task_32" = list(
    "model" = "rf",
    "feature_set" = "plasma_ab_titer+metadata+baseline",
    "table_col" = "3.2) CCL3-D3-FC-Rank"
  ),
  "task_41" = list(
    "model" = "lasso",
    "feature_set" = "plasma_ab_titer_pca+plasma_cytokine_concentration_by_legendplex_pca+t_cell_activation_pca+metadata+baseline",
    "table_col" = "4.1) IFNG/IL5-Polarization-D30-Rank"
  )
)

# checks
## 0) all task column names are in submission table
purrr::map_lgl(prediction_specs, ~ .x[["table_col"]] %in% colnames(submission_df)) %>%
  all() %>%
  stopifnot()

## a) all tasks are specified
stopifnot(dplyr::setequal(names(task_meta), names(prediction_specs)))

## b) all features are present
purrr::map_lgl(prediction_specs, ~ length(str_split(.x[["feature_set"]], "\\+")[[1]]) >= 1) %>%
  all() %>%
  stopifnot()

purrr::map_lgl(prediction_specs, ~ .x[["feature_set"]] %in% c(names(feature_list), "baseline")) %>%
  all() %>%
  stopifnot()

## c) which features are never selected?
selected_features <- purrr::map(prediction_specs, ~ str_split(.x[["feature_set"]], "\\+")[[1]]) %>%
  unlist() %>%
  unique()

all_features <- purrr::map(names(feature_list), ~ str_split(.x, "\\+")[[1]]) %>%
  unlist() %>%
  unique()

all_features[!all_features %in% selected_features]
rm(selected_features, all_features)

## d) which features are most commonly used
purrr::map(prediction_specs, ~ str_split(.x[["feature_set"]], "\\+")[[1]]) %>%
  unlist() %>%
  stringr::str_remove_all(., "_pca$") %>%
  table() %>%
  sort(decreasing=TRUE)

## e) all models are implemented
purrr::map_lgl(prediction_specs, ~ .x[["model"]] %in% c("rf", "rf+boruta", "elnet", "lasso", "baseline")) %>%
  all() %>%
  stopifnot()

## f) pull scores from result df
purrr::imap(prediction_specs, function(task_specs, task) {
  entry <- model_selection_df %>%
    dplyr::filter(task == !!task) %>%
    dplyr::filter(model == !!task_specs[["model"]]) %>%
    dplyr::filter(feature_set == !!task_specs[["feature_set"]])
  stopifnot(nrow(entry)==1)
  return(entry)
}) %>%
  dplyr::bind_rows()
```

3. Make the predictions

- Train model on full dataset

- Make predictions for the 2023 data

```{r}
var_imp_list <- list()
predictions_list <- list()

for (task_name in names(prediction_specs)) {
  # task_name <- "task_12"
  task_specs <- prediction_specs[[task_name]]
  
  print(task_name)
  print(task_specs$model)
  print(task_specs$feature_set)
  
  task_df <- target_list[[task_name]]
    
  target_df <- task_df %>%
    dplyr::select(subject_id, target)
  
  baseline_df <- baseline_list[[task_name]]
  
  if (task_specs$feature_set=="baseline" & task_specs$model=="baseline") {
    feature_df <- baseline_df %>% dplyr::select(subject_id)
  } else {
    feature_df <- feature_list[[task_specs$feature_set]]
  }
  
  if (str_detect(task_specs$feature_set, "baseline")) {
    feature_df <- feature_df %>%
      dplyr::left_join(baseline_df, by="subject_id")
  }
  
  train_df <- target_df %>%
    tidylog::inner_join(feature_df, by="subject_id") %>%
    dplyr::left_join(dplyr::distinct(dplyr::select(meta_data, c(subject_id, dataset))),
                     by="subject_id") %>%
    dplyr::select(-c(dataset)) %>%
    tibble::column_to_rownames(var="subject_id")
  stopifnot(!any(is.na(train_df)))
  stopifnot(nrow(train_df)==nrow(dplyr::distinct(train_df)))
  # check that features were duplicated during joining of tables (i.e. features end with .x and .y) -> but doesn't make sense with the PCA joining...
  #stopifnot(all(purrr::map_dbl(stringr::str_split(colnames(train_df), "\\."), ~ length(.x)) == 1))
  #colnames(train_df)[purrr::map_dbl(stringr::str_split(colnames(train_df), "\\."), ~ length(.x)) > 1]
  
  test_df <- submission_df %>%
    dplyr::mutate(subject_id = SubjectID) %>% 
    dplyr::select(subject_id) %>%
    dplyr::left_join(feature_df, by="subject_id") %>%
    tibble::column_to_rownames(var="subject_id")
  
  # check if the baseline is missing for certain instances
  if (str_detect(task_specs$feature_set, "baseline")) {
    if (sum(is.na(test_df$baseline)) > 0) {
      message(paste0("Baseline values are missing for ", 
                   sum(is.na(test_df$baseline)),
                   " samples! Imputing with Median"))
      test_df$baseline[is.na(test_df$baseline)] <- median(test_df$baseline, na.rm=TRUE)
    }
  }
  
  # now there should be no more NAs!
  stopifnot(mean(is.na(test_df)) == 0)
  
  # some more checks
  stopifnot(identical(colnames(train_df)[colnames(train_df)!="target"], colnames(test_df)))
  #colnames(test_df)[!colnames(test_df) %in% colnames(train_df)]
  #colnames(train_df)[!colnames(train_df) %in% colnames(test_df)]
  
  # if we need matrix, vector objects for certain ML algos
  train_mtx <- as.matrix(dplyr::select(train_df, -target))
  train_target <- dplyr::pull(train_df, target)
  test_mtx <- as.matrix(test_df)
  
  if (task_specs$feature_set=="baseline" & task_specs$model=="baseline") {
    predictions <- test_df$baseline
    names(predictions) <- rownames(test_df)
  } else if (task_specs$model == "rf") {
    model <- ranger::ranger(formula = target ~ ., 
                            data = train_df, 
                            num.trees = 500, 
                            importance = "impurity")
    predictions <- predict(model, test_df)$predictions
    names(predictions) <- rownames(test_df)
    var_imp_list[[task_name]] <- 
      tibble(task=task_name,
             importance = model$variable.importance,
             feature = names(model$variable.importance)
      )
  } else if (task_specs$model == "rf+boruta") {
    boruta_out <- Boruta::Boruta(formula = target ~ .,
                                 data = train_df,
                                 maxRuns=1000,
                                 num.trees=500 
                                 )
    boruta_feats <- names(boruta_out$finalDecision)[boruta_out$finalDecision=="Confirmed"]
    stopifnot(length(boruta_feats) != 0)
  
    train_df_subset <- train_df %>% dplyr::select(dplyr::all_of(c(boruta_feats, "target")))
    test_df_subset <- test_df %>% dplyr::select(dplyr::all_of(c(boruta_feats, "target")))
    model <- ranger::ranger(formula = target ~ ., 
                            data = train_df_subset, 
                            num.trees = 500, 
                            importance = "impurity")
    predictions <- predict(model, test_df_subset)$predictions
    names(predictions) <- rownames(test_df)
    var_imp_list[[task_name]] <- 
      tibble(task=task_name,
             importance = model$variable.importance,
             feature = names(model$variable.importance)
      )
  } else if (task_specs$model == "lasso") {
    # standardize the features, even though glmnet is doing this automatically!
    # feature_means <- matrixStats::colMeans2(train_mtx)
    # feature_sds <- matrixStats::colSds(train_mtx)
    # train_mtx <- t((t(train_mtx) - feature_means) / feature_sds)
    # train_target <- (train_target - mean(train_target)) / sd(train_target)
    #test_mtx <- t((t(test_mtx) - feature_means) / feature_sds)
    
    # inner hyperparameter optim loop using LOOCV to find best lambda
    model <- glmnet::cv.glmnet(x=train_mtx,
                               y=train_target,
                               nfolds=nrow(train_mtx),
                               grouped=FALSE,
                               alpha=1, # LASSO
                               type.measure="mae", # not sure whether this is better
                               standardize=TRUE, # by default each feature is standardized internally
                               family="gaussian"
                               )
    # investigate
    #beta_mtx <- as.data.frame(as.matrix(model$glmnet.fit$beta))
    #colnames(beta_mtx) <- paste0("lambda_", round(model$lambda, 3))
    #plot(model)
    #lambda_value <- model$lambda.1se, # model$lambda.min
    # never use zero features (with lamba=model$lambda[1]), since constant predictions
    # will have NA spearman...
    lambda_value <- ifelse(model$lambda.min==model$lambda[1], 
                           model$lambda[2], model$lambda.min)
    predictions <- predict(model,
                           newx=test_mtx, 
                           s=lambda_value)[,1]
    coefs <- model$glmnet.fit$beta[, lambda_value==model$glmnet.fit$lambda]
    var_imp_list[[task_name]] <- 
      tibble(task=task_name,
             importance = abs(coefs),
             feature = names(coefs)
      )
  } else if (task_specs$model == "elnet") {
    # standardize the features, even though glmnet is doing this automatically!
    # feature_means <- matrixStats::colMeans2(train_mtx)
    # feature_sds <- matrixStats::colSds(train_mtx)
    # train_mtx <- t((t(train_mtx) - feature_means) / feature_sds)
    # train_target <- (train_target - mean(train_target)) / sd(train_target)
    #test_mtx <- t((t(test_mtx) - feature_means) / feature_sds)
    
    # inner hyperparameter optim loop using LOOCV to find best lambda
    model <- glmnet::cv.glmnet(x=train_mtx,
                               y=train_target,
                               nfolds=nrow(train_mtx),
                               grouped=FALSE,
                               alpha=0.5, # elastic net
                               type.measure="mae", # not sure whether this is better
                               standardize=TRUE, # by default each feature is standardized internally
                               family="gaussian"
                               )
    lambda_value <- ifelse(model$lambda.min==model$lambda[1], 
                           model$lambda[2], model$lambda.min)
    predictions <- predict(model,
                           newx=test_mtx, 
                           s=lambda_value)[,1]
    coefs <- model$glmnet.fit$beta[, lambda_value==model$glmnet.fit$lambda]
    var_imp_list[[task_name]] <- 
      tibble(task=task_name,
             importance = abs(coefs),
             feature = names(coefs)
      )
  }
  
  stopifnot(all(names(predictions) == rownames(test_df)))
  
  predictions_list[[task_name]] <- 
    tibble::tibble(subject_id = as.numeric(names(predictions)), 
                   predictions = predictions,
                   task_name = task_name,
                   model = task_specs$model,
                   feature_set = task_specs$feature_set)
}
```

```{r}
names(predictions_list)
```

```{r}
submission_df_with_predictions <- submission_df

for (task_name in names(predictions_list)) {
  task_preds_df <- predictions_list[[task_name]] %>%
    dplyr::select(subject_id, predictions)
  
  submission_df_with_predictions <- submission_df_with_predictions %>%
    tidylog::left_join(task_preds_df, by=c("SubjectID" = "subject_id")) %>%
    dplyr::mutate(!!prediction_specs[[task_name]][["table_col"]] := predictions) %>%
    dplyr::select(-predictions)
}
submission_df_with_predictions
```

Lastly we need to create ranks from those predictions, see:

- https://discuss.cmi-pb.org/t/ranking-results-clarification/321

- We expect contestants to generate computational models, and upon making predictions, these values are ranked from highest to lowest (i.e. highest = 1, lowest = N) before making a final submission.

```{r}
submission_df_with_rank <- submission_df_with_predictions

task_titles <- purrr::map_chr(prediction_specs, ~ .x[["table_col"]])
stopifnot(all(task_titles %in% colnames(submission_df_with_predictions)))

for (task_title_oi in task_titles) {
  submission_df_with_rank[[task_title_oi]] <- 
    rank(submission_df_with_predictions[[task_title_oi]], ties.method="average")
}
readr::write_tsv(
  x = submission_df_with_rank, 
  file = file.path(output_dir, paste0(format(Sys.Date(), "%Y-%m-%d"), "_submission_psls.tsv"))
)
submission_df_with_rank
```

Then we can also look at the most important features

```{r}
var_imp_list %>%
  dplyr::bind_rows() %>%
  dplyr::group_by(task) %>%
  dplyr::slice_max(order_by=importance, n=10, with_ties = FALSE) %>%
  dplyr::filter(importance > 0)
  
var_imp_list %>%
  dplyr::bind_rows() %>%
  dplyr::group_by(task) %>%
  dplyr::slice_max(order_by=importance, n=10, with_ties = FALSE) %>%
  dplyr::filter(importance > 0) %>%
  dplyr::summarise(features = paste0(feature, collapse=", ")) %>%
  readr::write_delim(., file=file.path(output_dir, paste0(format(Sys.Date(), "%Y-%m-%d"), "_var_imp_df.tsv")), delim="\t")
```

# Appendix

Also write the image just in case

```{r}
save.image(file=file.path(output_dir, paste0(format(Sys.Date(), "%Y-%m-%d"), "_model_selection_session.RData")))
#load(file=file.path(output_dir, "2024-11-11_model_selection_session.RData"))
```

